Supplemental Materials for ``Universals of word order reflect optimization of grammars for efficient communication''

### 2019-10-12 writeup/pnas-word-order-si.tex
code

TODO CITE \cite{mielke-et-al-2019} % However,  a lan-guage’s vocabulary size and the length in charactersof its sentences were well-correlated with difficultyon our large set of languages.
\cite{gerz2018language}
% TODO maybe also some recent stuff in data-driven induction of typology https://www.mitpressjournals.org/doi/abs/10.1162/COLI_a_00357

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage.pdf}
\caption[Predictability and Parseability]{Predictability and parseability of 51 languages. Green: random baselines, Red: real grammar, blue: approximate Pareto frontier, computed from the optimized grammars.\footnotemark All data are $z$-scored. TODO adapt for colorblindness.}\label{fig:pareto-per-lang}
\end{figure}

%\begin{figure}
%\centering
%\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage.pdf}
%\caption[Predictability and Parseability]{Predictability and parseability of 51 languages. Green: random baselines, Red: real grammar, blue: approximate Pareto frontier, computed from the optimized grammars.\footnotemark All data are $z$-scored. TODO adapt for colorblindness.}\label{fig:pareto-per-lang}
%\end{figure}

@@

\begin{table}
\small{
\begin{center}
\begin{tabular}{|l||l|lll|llll|ll|llllll}
\hline
	& Mean $\beta$ & SD & Lower 95\% CrI & Upper 95\% CrI \\
\hline\hline
	Efficiency ($\lambda=0.9$) &	-5.88   &   1.08  &  -8.28  &  -3.97  \\
	Surp & -3.48   &   0.88  &  -5.42  &  -1.85   \\
	Pars & -5.55   &   1.08  &  -7.80  &  -3.67    \\
\hline
\end{tabular}
\end{center}
}
	\caption{TODO}\label{tab:study1}
\end{table}
% paretoByFamily.R

@@

	\caption{TODO For $\lambda=0$ and $\lambda=0.9$, we provide lines connecting points that have the same efficiency value as the real language (red cross). Points to the bottom right of the lines have lower efficiency under the respective values of $\lambda$. In this example, all baseline samples (green dots) have lower efficiency than the real language, even though the real language does not outperform every baselines on both components of efficiency.}\label{fig:opt-lambda-explanation}
\end{figure}

@@

TODO discuss difference from semantic typology:
TODO mention that, in semantic typology, random baselines might tend to be closer to the frontier, whereas, in our study, there tends to be a low-density region between the baseline grammars and the optimum (as also observed in previous DLM and surprisal studies, cite Futrell and Gildea/Jaeger).

@@

\paragraph{Bayesian Mixed-Effects Regression}
We modeled the probabilities $p_{L,j}$ that a grammar optimized for data from language $L$ satisfies the $j$-th correlation ($j=1,...,8$) using a multilevel logistic model \cite{gelman2013bayesian}, with random intercepts for the language for whose data the grammar had been optimized, and for its language family, annotated according to \url{http://universaldependencies.org/}.
Formally,
\begin{equation}\label{eq:mixed-effects}
logit(p_{L,j}) = \alpha_j + u_{L,j} + v_{f_L,j}
\end{equation}
where $f_L$ is the language family of $L$.
The intercepts $\alpha_j$ ($j=1,...8$) encode the population-level prevalence of the correlations when controlling for differences between datasets from different languages and language families; $u_{L,j}$, $v_{f_L,j}$ encode per-language and per-family deviations from the population-level intercept $\alpha_j$.
%, and
%$u_{L,1,...8} \sim \mathcal{N}(0, \Sigma_{lang})$,
%$v_{f_L,1,...8} \sim \mathcal{N}(0, \Sigma_{fam})$.

%As we constructed 4 verb-object and 4 object-verb grammars for each of the 51 languages, TODO...

@@

TODO also mention: As we estimate parseability and predictability using cross-entropies, our models constitute variational approximations to the true parseability and predictability values, similar to previous work quantifying other aspects of linguistic complexity \citep{cotterell2019complexity}.
TODO

@@

Prior work has suggested \emph{Dependency Length Minimization} (DLM) as a characteristic of efficient word order~\cite{futrell2015largescale,liu2017dependency,temperley2018minimizing}.
This is the idea that word order minimizes the average distance between syntactically related words.
It is known that human languages reduce dependency length compared to random baselines~\cite{futrell2015largescale,liu2017dependency,temperley2018minimizing}.
Prior work has suggested principles akin to DLM as approximating efficiency optimization of grammars~\cite{hawkins1994performance,hawkins2004efficiency,futrell2017memory, futrell2017generalizing}.
It is a heuristic formalization of the idea that long dependencies should create high memory requirements in parsing and prediction~\cite{hawkins1994performance,gibson1998linguistic,gibson2000dependency, futrell2017memory}.
Indeed, \cite{futrell2017memory} argues specifically that it emerges from efficiency optimization.
[TODO also cite rijkhof, ferrer i cancho]

@@

We asked whether efficiency optimization predicts dependency length minimization effects.
We first computed dependency length for grammars optimized for efficiency.
We found that 100\% of grammars optimized for efficiency reduce average dependency length compared to baseline grammars ($p < 0.05$, by one-sided $t$-test). 
This suggests that the reduction of dependency length observed in natural language is indeed predicted by efficiency maximization, confirming theoretical arguments made in prior work~\cite{hawkins1994performance,hawkins2004efficiency,futrell2017memory, futrell2017generalizing}.
Next, we constructed grammars that minimize average dependency length, using the same gradient descent method as we used for efficiency optimization (Section~\ref{sec:optim-eff}).
We expect that such grammars should have shorter dependency length than the real grammars, or grammars optimized for efficiency.
In Figure~\ref{fig:dlm-avg}, we plot the mean dependency length for optimized, real, and baseline orderings.
We find that optimizing grammars for efficiency reduces dependency length to a similar degree as found in the actual orderings in the corpora, almost up to the limit given by directly optimizing for dependency length. [TODO acknowledge  Ferrer i Cancho R, Liu H (2014) The risks of mixing dependency lengths from se-quences of different length.Glottotheory5(2):143–155.]
We also plot more detailed results for four languages in Figure~\ref{fig:dlm-4langs}, plotting dependency length as a function of sentence length as reported in prior work \cite{futrell2015largescale}.
Optiziming grammars for efficiency produces dependency lengths similar to those found in the actual orderings.

@@

>>>>

%TODO CITE \cite{mielke-et-al-2019} % However,  a lan-guage’s vocabulary size and the length in charactersof its sentences were well-correlated with difficultyon our large set of languages.
%\cite{gerz2018language}
% TODO maybe also some recent stuff in data-driven induction of typology https://www.mitpressjournals.org/doi/abs/10.1162/COLI_a_00357

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage.pdf}
\caption[Predictability and Parseability]{Predictability and parseability of 51 languages. Green: random baselines, Red: real grammar, blue: approximate Pareto frontier, computed from the optimized grammars.\footnotemark All data are $z$-scored.}\label{fig:pareto-per-lang}
\end{figure}

@@

\begin{table}
\small{
\begin{center}
\begin{tabular}{|l||l|lll|llll|ll|llllll}
\hline
	& Mean $\beta$ & SD & Lower 95\% CrI & Upper 95\% CrI \\
\hline\hline
	Efficiency ($\lambda=0.9$) &	-5.88   &   1.08  &  -8.28  &  -3.97  \\
	Surp & -3.48   &   0.88  &  -5.42  &  -1.85   \\
	Pars & -5.55   &   1.08  &  -7.80  &  -3.67    \\
\hline
\end{tabular}
\end{center}
}
	\caption{Models estimating the log-odds of a random baseline grammar improving over a real grammar in efficiency ($\lambda=0.9$), surprisal, or parseability, with random effects for languages and language families. The strongly negative estimates of $\beta$ confirm that, across languages and language families, real grammars improve over most baselines. TODO}\label{tab:study1-glmer}
\end{table}

@@

	\caption{For $\lambda=0$ and $\lambda=0.9$, we provide lines connecting points that have the same efficiency value as the real language (red cross). Points to the bottom right of the lines have lower efficiency under the respective values of $\lambda$. In this example, all baseline samples (green dots) have lower efficiency than the real language, even though the real language does not outperform every baselines on both components of efficiency.}\label{fig:opt-lambda-explanation}
\end{figure}

@@

TODO
We note that not all real grammars are on the estimated Pareto frontier.
TODO discuss difference from semantic typology:
TODO mention that, in semantic typology, random baselines might tend to be closer to the frontier, whereas, in our study, there tends to be a low-density region between the baseline grammars and the optimum (as also observed in previous DLM and surprisal studies, cite Futrell and Gildea/Jaeger).

@@

\paragraph{Bayesian Mixed-Effects Regression}
We modeled the probabilities $p_{L,j}$ that a grammar optimized for data from language $L$ satisfies the $j$-th correlation ($j=1,...,8$) using a multilevel logistic model \cite{gelman2013bayesian}, with random intercepts for the language for whose data the grammar had been optimized, and for its language family, annotated according to \url{http://universaldependencies.org/}.
Formally,
\begin{equation}\label{eq:mixed-effects}
logit(p_{L,j}) = \alpha_j + u_{L,j} + v_{f_L,j}
\end{equation}
where $f_L$ is the language family of $L$.
The intercepts $\alpha_j$ ($j=1,...8$) encode the population-level prevalence of the correlations when controlling for differences between datasets from different languages and language families; $u_{L,j}$, $v_{f_L,j}$ encode per-language and per-family deviations from the population-level intercept $\alpha_j$.

@@

%TODO also mention: As we estimate parseability and predictability using cross-entropies, our models constitute variational approximations to the true parseability and predictability values, similar to previous work quantifying other aspects of linguistic complexity \citep{cotterell2019complexity}.
%TODO

@@

Prior work has suggested \emph{Dependency Length Minimization} (DLM) as a characteristic of efficient word order~\cite{futrell2015largescale,ferrericancho2004euclidean,liu2017dependency,temperley2018minimizing}.
This is the idea that word order minimizes the average distance between syntactically related words.
It is known that human languages reduce dependency length compared to random baselines~\cite{liu2010dependency,futrell2015largescale,liu2017dependency,temperley2018minimizing}.
Prior work has suggested principles akin to DLM as approximating efficiency optimization of grammars~\cite{hawkins1994performance,hawkins2004efficiency,futrell2017memory, futrell2017generalizing}.
It is a heuristic formalization of the idea that long dependencies should create high memory requirements in parsing and prediction~\cite{hawkins1994performance,gibson1998linguistic,gibson2000dependency, futrell2017memory}.
Indeed, \cite{futrell2017memory} argues specifically that it emerges from efficiency optimization.
%[TODO also cite rijkhof, ferrer i cancho]

@@

We asked whether efficiency optimization predicts dependency length minimization effects.
We first computed dependency length for grammars optimized for efficiency.
We found that 100\% of grammars optimized for efficiency reduce average dependency length compared to baseline grammars ($p < 0.05$, by one-sided $t$-test). 
This suggests that the reduction of dependency length observed in natural language is indeed predicted by efficiency maximization, confirming theoretical arguments made in prior work~\cite{rijkhoff1986word,rijkhoff1990explaining,hawkins1994performance,hawkins2004efficiency,futrell2017memory, futrell2017generalizing}.
Next, we constructed grammars that minimize average dependency length, using the same gradient descent method as we used for efficiency optimization (Section~\ref{sec:optim-eff}).
We expect that such grammars should have shorter dependency length than the real grammars, or grammars optimized for efficiency.
In Figure~\ref{fig:dlm-avg}, we plot the mean dependency length for optimized, real, and baseline orderings.
We find that optimizing grammars for efficiency reduces dependency length to a similar degree as found in the actual orderings in the corpora, almost up to the limit given by directly optimizing for dependency length. %[TODO acknowledge  Ferrer i Cancho R, Liu H (2014) The risks of mixing dependency lengths from se-quences of different length.Glottotheory5(2):143–155.]
We also plot more detailed results for four languages in Figure~\ref{fig:dlm-4langs}, plotting dependency length as a function of sentence length as reported in prior work \cite{futrell2015largescale}.
Optiziming grammars for efficiency produces dependency lengths similar to those found in the actual orderings.

@@

% paretoByFamily.R
